\sys: Controllable Program Regeneration

### 2025-07-31 docs/plos/draft.tex
Redesign section names
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>
Many fixes
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\promptbox}[1]{%
  \vspace{0.5em}%
  \hspace*{-0.0135\columnwidth}%
  \fbox{%
    \parbox[t]{0.95\columnwidth}{\tt #1}%
  }%
  \vspace{0.5em} % adjust this value as desired
}

@@

\setlength{\intextsep}{0pt} % vertical space above/below
\setlength{\columnsep}{5pt} % horizontal space between text and figure

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis and are checked post hoc to ensure
  correctness.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}

@@

\begin{figure}[t]
  % https://docs.google.com/drawings/d/1ozMu2B0RUhyLWC_rM6GLBrioov80-M06CBqYRAhCk18/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_projection.pdf}
  \caption{\textbf{Property projection and verifiable regeneration in \sys.}
  Instead of transforming source code directly, \sys projects the original
  program into a minimal set of verifiable properties---such as input/output
  examples, system call traces, \etc.
  These properties serve as declarative
  constraints for synthesis.
  The regenerated implementation is then checked to
  ensure it satisfies the specified properties, providing a verifiable
  foundation for trust unlike purely prompt-based LLM workflows.}
  \label{fig:projection}
\end{figure}

@@

Recent advances in natural-language processing, particularly large language
models (LLMs) have opened new possibilities for automating program
transformation and synthesis tasks,
making them attractive tools for software
evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for an that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating
third-party modules that must be adapted to local conventions or hardened
against security vulnerabilities.
For example, a developer can regenerate a C library as idiomatic
JavaScript while preserving I/O equivalence, or decompose a monolithic function
into modular components suitable for sandboxing.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys can function not only as a standalone tool but also as a
library, where other components can invoke it as a programmable,
constraint-guided regeneration module.

@@

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively across a codebase.
  Edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied as part of the example~(\cref{sec:example}).
  Dotted edges signify ommited layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

@@

After 30.4 seconds and one attempt, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

Building on the same source as the previous example, \sys can perform
cross-language translation.
In this scenario, the developer aims to lift input
handling logic into a sandboxed JavaScript runtime to allow controlled
developer hooks and runtime customization, while reducing the attack surface of
the C runtime.
By moving this function into the scripting environment, it
becomes easier to enforce memory safety and auditability constraints.

\begin{wrapfigure}[5]{r}{.46\columnwidth}
\begin{minted}{prolog}
language(p_, javascript).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
llm_translate(Ic, Ijs),
llm_translate(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query uses \texttt{llm\_translate}
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: $\{\texttt{graph(P, I, O)}\}$.
It then invokes \texttt{llm\_translate} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.55\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

The resulting modular program is:
\begin{minted}{javascript}
function connectDB() {
  const sqlite = require('sqlite');
  return sqlite.open({filename: 'game.db'}); }

@@

\sys is a programming framework for program regeneration. It enables developers to express transformation goals declaratively—as constraints over the desired properties of regenerated code—and automatically synthesizes implementations that satisfy those constraints.

At a high level, \sys operates in three stages:

\begin{enumerate}
  \item \textbf{Property Specification and Extraction}: Users define goals using logical constraints over program properties (e.g., language, side-effect freedom, input/output equivalence). Plugins contribute analyses and extraction routines to populate these properties from the input program.
  \item \textbf{Constraint-Guided Synthesis}: The framework converts property constraints into structured prompts for a language model. These prompts describe what must be preserved or changed, including examples and declarative requirements.
  \item \textbf{Verification and Feedback}: Synthesized candidates are validated against the target properties. Failures trigger retries or incremental refinements until the goals are satisfied or search is exhausted.
\end{enumerate}

This design makes regeneration tasks programmable, composable, and extensible, without requiring developers to encode transformation logic manually.

\section{Discussion}
\label{sec:evaluation}

We evaluate \sys across several dimensions: its ability to regenerate correct and maintainable code, the impact of property-driven guidance, and practical considerations around applying it to diverse inputs. This evaluation is preliminary and intended to illustrate feasibility and trade-offs.

\subsection{Benchmarks}

Our benchmark suite comprises 37 tasks spanning simple translation, obfuscation removal, and security hardening (Table~\ref{tab:benchmarks}). Inputs were drawn from Rosetta Code, popular npm and PyPi packages, known supply-chain attacks, and selected obfuscated programs from the IOCCC.

A regeneration is counted as successful if the output satisfies the declared properties and passes either functional tests or manual inspection.

\begin{table}[ht]
\centering
\caption{\textbf{Benchmark Overview}.}
\begin{tabular}{lrll}
\toprule
Benchmark & N & Purpose & Sources \\
\midrule
Rosetta Code & 16 & Cross-language translation & \cite{rosettacode} \\
Short Utilities & 13 & Parity preservation & \cite{regbench2025} \\
SSCA & 3 & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC & 2 & Deobfuscation & \cite{ioccc} \\
% Microbenchmarks & 3 & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

\subsection{Correctness and Functional Equivalence}

Overall, \sys regenerated correct code in most cases. All Rosetta Code and utility tasks were completed successfully. For supply-chain attacks, \sys removed malicious behaviors while preserving legitimate functionality, confirming that treating the original source as untrusted can be effective.

@@

\subsection{Impact of Property-Guided Synthesis}

A key observation is that constraining the synthesizer to operate only over extracted properties significantly improved reliability. For example, when regenerating modules with hidden side effects, naive prompting often reproduced those side effects verbatim. By contrast, \sys’s selective extraction of I/O traces and signatures avoided this overfitting.

Prompt size was also reduced substantially (up to 60\%), and synthesis iterations converged faster.

\subsection{Comparison to Naive Prompting}

As a baseline, we prompted a large language model with the full source code and a natural language description of the task. For simple translation and trivial refactoring, this approach often succeeded. However, it frequently failed in cases requiring property preservation without leaking implementation details.

@@

\section{Related Work}

\heading{Program synthesis}
Program synthesis research spans a rich space of techniques.
Classical techniques emphasize correctness and
interpretability, synthesizing programs from specifications~\cite{alur2013syntax, feser2015synthesizing, gulwani2011automating,leino2016dafny},
type or syntax constraints \cite{polikarpova2016program,reynolds2019syguscomp},
or examples~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming},
which are provided either by users or automatically inferred~\cite{cambronero2019active,harp:ccs:2021}
These works often target narrowly
defined domains such as string manipulation~\cite{harp:ccs:2021} or data migration
\cite{yaghmazadeh2018automated} and focus on ensuring provable guarantees.
More recently, LLM-based methods~\cite{austin2021program, chen2021evaluating}
explore broad-domain code generation via prompt-based conditioning.
While flexible, these models are prone to hallucination and struggle
with reasoning under negation or satisfying structured constraints~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode,jiang2024llmsdreamelephantswhen,hwang2024thinkpinkelephant}.

\sys leverages insights from both these paradigms.
Unlike purely symbolic or purely neural approaches, it leverages LLMs within a
constrained planning framework to guide transformations while remaining
responsive to property-based requirements.

\heading{Program transformations and refactoring}
The evolution and maintenance of software systems often involve transformations
such as refactoring~\cite{Fowler99,Mens04,Myers16}, translation, or security hardening. % TODO: Add more citations

Tools in this space often operate via syntax-tree transformations or
domain-specific templates. Verification-based methods such as \textsc{Dafny}
\cite{leino2016dafny} and \textsc{CBMC} \cite{Clarke04} ensure functional correctness
but require heavy formalization.
Others automate migration via
programming-by-example or interactive guidance \cite{gulwani2017program, le2017interactive}.

\sys aims to generalize this landscape by treating transformations as
goal-directed regenerations---rather than prescriptive rewrites, driven by
properties such as program behaviors and architectural goals, expressed declaratively.

\heading{Logic programming and constraint solving}
Logic programming has long served as a foundation for reasoning in programming
tools, offering a declarative model for expressing properties and constraints.
Answer Set Programming (ASP), in particular, provides a rich framework for
expressing defaults, exceptions, and optimization via stable model semantics
\cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
Paradigms like ASP have been used for
declarative program analysis \cite{benton2007interactive}, automated planning
\cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
representation in verification and synthesis pipelines.

>>>>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\rtype}[1]{\setlength{\fboxsep}{1.75pt}\colorbox{lightperiwinkle}{\texttt{#1}}}
\newcommand{\promptbox}[1]{%
  \vspace{0.5em}%
  \hspace*{-0.0135\columnwidth}%
  \fbox{%
    \parbox[t]{0.95\columnwidth}{\tt #1}%
  }%
  \vspace{0.5em} % adjust this value as desired
}

@@

\setlength{\intextsep}{0pt} % vertical space above/below
% \setlength{\columnsep}{5pt} % horizontal space between text and figure

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are common-place, but as brittle and time-consuming as ever.
  Towards their automation, large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
  This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  the program's input/output behavior, execution traces, or interaction with external databases.
  These properties are extracted and serve the dual purpose of guiding synthesis and guardrailing synthesized outputs.
  \sys unifies established techniques and inspires new ones. % TODO: Rework this
  A preliminary evaluation shows that \sys succeeds even on tasks where state-of-the-art LLMs fail.
\end{abstract}

@@

\begin{figure}[t]
  % https://docs.google.com/drawings/d/1ozMu2B0RUhyLWC_rM6GLBrioov80-M06CBqYRAhCk18/edit
  \includegraphics[width=.7\columnwidth]{figs/kvasir_projection.pdf}
  \caption{\textbf{Property projection and verifiable regeneration in \sys.}
  Instead of transforming source code directly, \sys projects the original
  program into a minimal set of verifiable properties---such as input/output
  examples, system call traces, \etc.
  The regenerated implementation is then checked to
  ensure it satisfies the specified properties, providing a verifiable
  foundation for trust unlike purely prompt-based LLM workflows.}
  \label{fig:projection}
\end{figure}

@@

Recent advances in natural-language processing, particularly large language
models (LLMs) have opened new possibilities for automating program
transformation and synthesis tasks,
making them attractive tools for software
evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through ad-hoc heuristics, manual trial-and-error, or high-quality (and thus hard to create) examples~\cite{khattab2023dspy,brown2020languagemodelsfewshotlearners},
leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for an that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

\heading{Rethinking program transformations}
XXX Here, talk about bidirectional analysis, how now having a program that extracts 
a propoerty means that plugging it into sys will allow you to regenerate programs that hold or max/min-inimize over that property

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating
third-party modules that must be adapted to local conventions or hardened
against security vulnerabilities.
% For example, a developer can regenerate a C library as idiomatic
% JavaScript while preserving I/O equivalence, or decompose a monolithic function
% into modular components suitable for sandboxing.
% However, \sys is not designed for whole-program synthesis or large-scale migrations.
% Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
% To this end, \sys can function not only as a standalone tool but also as a
% library, where other components can invoke it as a programmable,
% constraint-guided regeneration module.

@@

\section{Point Solutions}
\label{sec:problem}

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow 
either through malice or by accident.
Then, thinking about a recently incorporated scripting component 
that allows players to make custom extensions to key parts of client-side game logic using JavaScript 
gives the idea tha maybe code that deals with user input should all be moved inside 
the sandboxed script engine.
Furthermore, the development team would like to identify functions that could 
be broken-down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{codeql, MORE} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target lagnuage.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes co-exist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the afformentioned trasnformations.
However, their problems X, Y, Z. 
There have been systems that provide ranging amounts of control of LLM-generated 
programs, but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strenghths 
and allows the programmer to steer their output and be aware of incorrect generations.

% % Related work that fails to address the problem
% \heading{Program synthesis}
% Program synthesis research spans a rich space of techniques.
% Classical techniques emphasize correctness and
% interpretability, synthesizing programs from specifications~\cite{alur2013syntax, feser2015synthesizing, gulwani2011automating,leino2016dafny},
% type or syntax constraints \cite{polikarpova2016program,reynolds2019syguscomp},
% or examples~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming},
% which are provided either by users or automatically inferred~\cite{cambronero2019active,harp:ccs:2021}
% These works often target narrowly
% defined domains such as string manipulation~\cite{harp:ccs:2021} or data migration
% \cite{yaghmazadeh2018automated} and focus on ensuring provable guarantees.
% More recently, LLM-based methods~\cite{austin2021program, chen2021evaluating}
% explore broad-domain code generation via prompt-based conditioning.
% While flexible, these models are prone to hallucination and struggle
% with reasoning under negation or satisfying structured constraints~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode,jiang2024llmsdreamelephantswhen,hwang2024thinkpinkelephant}.

% \sys leverages insights from both these paradigms.
% Unlike purely symbolic or purely neural approaches, it leverages LLMs within a
% constrained planning framework to guide transformations while remaining
% responsive to property-based requirements.

% \heading{Program transformations and refactoring}
% The evolution and maintenance of software systems often involve transformations
% such as refactoring~\cite{Fowler99,Mens04,Myers16}, translation, or security hardening. % TODO: Add more citations

% Tools in this space often operate via syntax-tree transformations or
% domain-specific templates. Verification-based methods such as \textsc{Dafny}
% \cite{leino2016dafny} and \textsc{CBMC} \cite{Clarke04} ensure functional correctness
% but require heavy formalization.
% Others automate migration via
% programming-by-example or interactive guidance \cite{gulwani2017program, le2017interactive}.

% \sys aims to generalize this landscape by treating transformations as
% goal-directed regenerations---rather than prescriptive rewrites, driven by
% properties such as program behaviors and architectural goals, expressed declaratively.

% \heading{Logic programming and constraint solving}
% Logic programming has long served as a foundation for reasoning in programming
% tools, offering a declarative model for expressing properties and constraints.
% Answer Set Programming (ASP), in particular, provides a rich framework for
% expressing defaults, exceptions, and optimization via stable model semantics
% \cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
% Paradigms like ASP have been used for
% declarative program analysis \cite{benton2007interactive}, automated planning
% \cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
% representation in verification and synthesis pipelines.

% \sys uses ASP as its planning back-end for the planning, decision-making
% process, while also uses its straightforward optimization capabilities.
% For each transformation task, \sys formulates the property-satisfaction problem as a logic query and uses an ASP solver to
% synthesize a feasible transformation plans.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively inside a codebase.
  In the above graph, edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied to~(\cref{sec:example}).
  Dotted edges signify ommited layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

@@

After 30.4 seconds and two attempts, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

\heading{Cross-language translation}
Building on the same source as the previous example, \sys can perform
cross-language translation.
In this scenario, the developer aims to lift input
handling logic into a sandboxed JavaScript runtime to allow controlled
developer hooks and runtime customization, while reducing the attack surface of
the C runtime.
By moving this function into the scripting environment, it
becomes easier to enforce memory safety and auditability constraints.

\begin{wrapfigure}[5]{r}{.46\columnwidth}
\begin{minted}{prolog}
language(p_, javascript).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
     tr(Ic, Ijs),
     tr(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query uses \texttt{llm\_translate}
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: $\{\texttt{graph(P, I, O)}\}$.
It then invokes \texttt{llm\_translate} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.56\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{graph(p)}
from the original program
the captures captures the input-output behavior on a generated set of
inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

The resulting program which now conists of two functions is:
\begin{minted}{javascript}
function connectDB() {
  const sqlite = require('sqlite');
  return sqlite.open({filename: 'game.db'}); }

@@

\begin{wrapfigure}{r}{.4\columnwidth}
  % https://docs.google.com/drawings/d/1d0piUrtExfTtUh_RuPfJFt1uZFOBEp0k-mHKSnogu-4/edit
  \includegraphics[width=.4\columnwidth]{figs/kvasir_levels.pdf}
  \label{fig:levels}
  \caption{The four key ingredients 
  of \sys that can be developed in parallel.}
\end{wrapfigure}
For \sys to enable developers
to express transformation goals declaratively---as constraints over the desired
properties of regenerated code---and automatically synthesizes implementations
that satisfy those constraints.

% The three key subsystems are:

% \begin{enumerate}
%   \item \textbf{Property Specification and Extraction}: Users define goals
%   using logical constraints over program properties (e.g., language,
% side-effect freedom, input/output equivalence). Plugins contribute analyses and
% extraction routines to populate these properties from the input program.
%   \item \textbf{Constraint-Guided Synthesis}: The framework converts property
%     constraints into structured prompts for a language model. These prompts
%     describe what must be preserved or changed, including examples and
%     declarative requirements.
%   \item \textbf{Verification and Feedback}:
%       Synthesized candidates are validated against the target properties.
%       Failures trigger retries or incremental refinements until the goals are
%       satisfied or search is exhausted.
% \end{enumerate}

% This design makes regeneration tasks programmable, composable, and extensible, without requiring developers to encode transformation logic manually.

\heading{Extensibility through plugins}
In order to maximize the applicability of the \sys framework 
towards many different kinds of transformations,
it is designed as an extensible platform rather than a monolithic tool.
All core functionality is provided by plugins, each contributing
the following functionality in the form three runtime hooks:
(1) \ttt{apply} which obtains static or dynamic properties from the original program (\eg
I/O traces, type signatures, or side-effect summaries),
(2) \ttt{verify} that will check the regenerated program against these properties (re-extracting them and comparing them to the original ones),
and
(3) \ttt{know} adds pre/post-conditions that inform regeneration planning.
The vision of \sys it to encourages community-driven growth, where specialized plugins---\eg for
security invariants or performance contracts---can coexist and interoperate
within a single regeneration framework~(\cref{sec:discussion}).

\heading{A declarative query language}
To detach itself from the details of the underlying analysis and

This separation of concerns---intent expressed declaratively, execution handled
automatically---makes regeneration auditable, repeatable, and free from brittle,
manual prompt-specific tuning.

\heading{Regeneration guardrails}
The final ingredient is a synthesis–verification loop that turns declarative goals into trustworthy code.
Given extracted properties and a query, Kvasir synthesizes candidate implementations using LLMs, leveraging their flexibility and knowledge of diverse programming idioms.
Each candidate is then checked against the declared properties: does it preserve the specified behavior, meet language or security constraints, and satisfy any additional invariants?
Failures trigger retries, refined prompts, or additional property extraction,
while successes provide an auditable proof of meeting the user’s intent,
at least as far as the properties allow (which is still a lot more than most LLM-based systems can guarantee).

\heading{Prototype implementation}
The \sys prototype implemetation uses Python vXXX,
the Pluggy library to define pluggable hooks, compiles queries to Clingo and uses the clasp
solver to arrive at a stable model which is converted to synthesis plans. 
XXX plugins have been implemented. All plugins range from XXX to XXX lines-of-code,
with the entire codebase being XXX lines-of-code.

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing—researchers rely on ad hoc examples that are hard to compare and often
oversimplified. This lack of a shared baseline obscures progress: when one
approach claims success on “language migration” and another on “security
hardening,” we have no way to tell whether they solve the same problem, let
alone which approach generalizes better.

Kvasir takes a step in this direction with its prototype benchmarks.
This suite contains tasks across several dimensions: its
ability to regenerate correct and maintainable
code, the impact of property-driven guidance,
and practical considerations around applying it
to diverse inputs. This evaluation is
preliminary and intended to illustrate
feasibility and trade-offs.

% TODO: Describe what a good benchmark suite would look like
The benchmark suite comprises 37 tasks spanning simple translation, obfuscation
removal, and security hardening (\cref{tab:benchmarks}). Inputs were drawn
from Rosetta Code, popular npm and PyPi packages, known supply-chain attacks,
and selected obfuscated programs from the IOCCC.

A regeneration is counted as successful if the output satisfies the declared
properties and passes both functional tests (when present) and manual inspection.

\begin{table}[t]
\centering
\caption{\textbf{Benchmark Overview}. 
The N column shows the number of programs/functions from each benchmark;
the Purpose column describes the regeneration task.}
\begin{tabular}{lrrll}
\toprule
Benchmark   & N  & Purpose                    & Sources \\
\midrule
RosettaCode & 16 & Cross-language translation & \cite{rosettacode} \\
Utilities   & 13 & Parity preservation        & \cite{regbench2025} \\
SSCA        & 3  & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC       & 2  & Deobfuscation              & \cite{ioccc} \\
% Microbenchmarks & 3  & LoC & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

\heading{Correctness results}
Overall, \sys regenerated correct code in most cases. All Rosetta Code and
utility tasks were completed successfully. For supply-chain attacks, \sys
removed malicious behaviors while preserving legitimate functionality,
confirming that treating the original source as untrusted can be effective.

@@

\heading{Comparison to naive prompting}
As a baseline, we prompted a large language model with the full source code and
a natural language description of the task. For simple translation and trivial
refactoring, this approach often succeeded. However, it frequently failed in
cases requiring property preservation without leaking implementation details.

@@

\section{More Challenges}
\label{sec:limitations}
